### 2021-08-10 reviewer_answers/Response to reviewers.tex

AUTHOR answer: Indeed, thanks for pointing this out to us. Section 3.3.4 has now been updated according to the above comment.

\item
Line 528 please, replace 'The figure 2 below' with 'Figure 2 below'

AUTHOR answer: This has been corrected.

\item
Your analysis is also built on the existence of statistical moments, hence in the case of a Cauchy distribution (which has a very 'normal' appearance), the model may lead to problems. It could be interesting to do a numerical experiment with a Cauchy distribution function. ($f(x) = 1/\pi \cdot 1 / (x^2 + 1)$ is the standard Cauchy distribution, which has no moments (average, standard deviation ...).

@@

\item
The errors in the figures, as well as the tolerances used in the paper (for instance $TOL_{res}$) seem to be large to me. Perhaps, they are not large relatively. The Authors should put this into its context, I think.

AUTHOR answer: We chose not to normalize the errors and estimators used in the numerics such that the a posteriori modeling-remainder estimator, $\eta_{rem,i}^k$ (shown in figures 6 and 13) is calculated exactly as given by Theorem 6.3, eq. (6.12d). The reason for such large values is that $\eta_{rem,i}^k$ is scaled by $G^{-\frac{1}{2}}$, where $G$ is the largest eigenvalue of the diffusion tensor $g$, which is chosen as a rather small constant for both examples (i.e., $5e-6$ and $2e-6$, respectively). We have added a sentence in Section 7, third paragraph where we comment on this.

\item
Line 768: Please specify which figures you are describing in terms of their labels rather than talking about 'The figures below'.

AUTHOR answer: This paragraph has been rewritten to make it more clear which figures are referred to. Similarly with the first paragraph of section (7.2.2).

@@

\begin{enumerate}
\item
The authors discuss and propose a model for the speciation processes leading to new species. To do so, they make use of a discretized version of the above equations. The discussions on the speciation processes seems to be highly related to the choice of the discrete system and in particular on the reconstruction operator. No connection with the original continuous system is clearly explained neither how it depends on the discretization.
Speciation is modeled in term of clustering in the trait space. Does it depend on the traits kept in the model?

AUTHOR answer:

\item
From a modelling view point, how do you choose the selection function $\alpha=\alpha(x,y,t)$? How is it related to the coefficients $A_{i,j}$ in (2.5a)? How the multi-scale parameters are coupled?

AUTHOR answer: $A_{i,j}$ is related to $\alpha$ through eq. (2.6c), i.e., the effect of species $i$ on species $j$ (as opposed to how individuals with different trait coordinates effect each other, as is the case at the population level). From a modeling point of view, $\alpha$ defines the ecological interactions in the system, and measures potentially competitive, mutualistic, and trophic interactions, and may in general be either symmetric or unsymmetric. Hence, it may take many different functional forms. For the numerical examples in the paper (which are only meant to illustrate the multi-scale framework, not represent any real eco-evolutionary system) we chose in the first case a symmetric competitive ecology, and in the second case an unsymmetric trophic ecology. More details on $\alpha$ can be found in [2]. However it is outside the scope of the paper to address in detail how $\alpha$ should be chosen in order to model different ecologies.

### 2021-08-13 reviewer_answers/Response to reviewers.tex

\item
The diffusion (second spatial derivative) term in the left-hand side of equation (2.2) represents the evolutionary process. I would call g the diffusion tensor instead of the diffusion coefficient (line 180).

AUTHOR answer: This has been corrected. We searched for other instances of "diffusion coefficient" aswell, but there were none.

\item
Since g is the diffusion tensor. It is assumed to be symmetric (Assumption 3.1). Of course, this symmetry facilitates the mathematical treatment with respect to well-posedness and error estimation. It is said that in the manuscript that Assumptions 3.1 follow from biology. Could there be any biological application cases where g is nonsymmetric? \\

AUTHOR answer: 
%Jeg tror ikke vi trenger symmetri, vi trenger sannsynligvis bare at den er positiv definit. Symmetri gir reelle egenverdier, og jeg har alltid tenkt at det var naturlig. Men naa som jeg reflekterer over det er det kanskje mulig (alt er mulig i biologi) med komplekse egenverdier, saa lenge den reelle parten er positiv.

@@

\item
Regarding the boundary condition that n vanishes on the boundary of the trait space (assumung that the trait space is 'large enough'), one could also relax this BC by considering g grad n + A n = 0. Although this BC would relax that n = 0, one has to fiddle around with the A-parameter. Do you think that this alternative BC could do a better job?

AUTHOR answer: 
%For finite domains, it is true that this is possible. I think it suffices to answer that we agree with the reviewer that this is conceptually an option, but we prefer to avoid introducing more (spatially dependent) parameter functions, and therefore retain the exposition with zero BCs.

@@

where $X \sim \mathcal{N}(0,\sigma^2)$. Of course, this relation is based on independence. Using the eigenvalues of the covariance (which is symmetric and hence the eigenvectors are orthogonal), the expression could nevertheless provide some estimation of the upper bound of the initial condition on the boundary.
Of course, this is not necessary to quantify directly for this manuscript, but the Authors could incorporate this in their future studies on this matter.

AUTHOR answer: We agree this is not necessary to quantify for the present purposes, but in any case thank you for pointing this out to us, it may very well become useful for future work on this topic.

\item
The errors in the figures, as well as the tolerances used in the paper (for instance $TOL_{res}$) seem to be large to me. Perhaps, they are not large relatively. The Authors should put this into its context, I think.

AUTHOR answer: We chose not to normalize the errors and estimators used in the numerics such that the a posteriori modeling-remainder estimator, $\eta_{rem,i}^k$ (shown in figures 6 and 13) is calculated exactly as given by Theorem 6.3, eq. (6.12d). The reason for such large values is that $\eta_{rem,i}^k$ is scaled by $G^{-\frac{1}{2}}$, where $G$ is the largest eigenvalue of the diffusion tensor $g$, which is chosen as a rather small constant for both examples (i.e., $5e-6$ and $2e-6$, respectively). We have added a discussion on this in Section 7.2.2, first paragraph (p. 20).

@@

AUTHOR answer: Thank you for the positive evaluation of our manuscrupt, we really appreciated your thoughtful comments!
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Referee \#2  (Remarks to the Author):}\\

This manuscript is concerned with a mathematical modelling for speciation using a general Lotka-Volterra equation modelling the evolution of some ecosystem. This model takes into account species traits and mutations through a diffusion term. This general equation is coupled with a system of ODE modelling the species level: density of species together with the evolution of the mean traits and the covariance matrix.

\begin{enumerate}
\item
The authors discuss and propose a model for the speciation processes leading to new species. To do so, they make use of a discretized version of the above equations. The discussions on the speciation processes seems to be highly related to the choice of the discrete system and in particular on the reconstruction operator. No connection with the original continuous system is clearly explained neither how it depends on the discretization.
Speciation is modeled in term of clustering in the trait space. Does it depend on the traits kept in the model?

AUTHOR answer: Indeed, there is missing a discussion on in what sense the time-discrete system relates to the original continuous one. To rectify this, we have rewritten the first paragraph of Section 3 (p. 7) where we elaborate on this.

%\red{Oi. Denne revieweren har ikke forstaatt noenting. Jeg foreslaar at vi hofflig svarer at vi bruker konvergent diskretiseringsmetode, og at kvaliteten paa rekonstruksjonsoperatoren styres av a posteriori feilestimatene.}

### 2021-08-13 manuscript/ex_article.tex
after talk with jan

Find $N :(0,T) \rightarrow \real^{s(1+d+d^2)}$ such that $n_i \geq 0$, for $1\leq i \leq s$, and 
\bse
\begin{alignat}{2}
\dfrac{\textnormal{d}N}{\textnormal{d}t} &= \Ncal(N), \quad && t\in(0,T), \label{ODEmodel}\\
N &= N_0, &&t = 0. \label{ODEmodelIC}
\end{alignat}
\ese
In practice, the problem \eqref{ODEmodel}--\eqref{ODEmodelIC} can be reduced to $s(1+d+d(d+1)/2)$ equations due to the symmetry of the trait covariance matrix. For additional details including a derivation of the above model, see~\cite{nordbotten2020dynamics}. Note also that when indexing the components of a point $x\in\Omega$ we use subscripts, i.e., $x = (x_1,\cdots,x_d)$, not to be confused with the species mean trait coordinate $x_i$, where we write $x_i(t) = (x_{i,1}(t),\cdots,x_{i,d}(t))$. 
%The distinction should be clear from the context.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relationship between population and species level models}
If there is a time interval $J := (t_{0},t_1) \subset (0,T)$ where $s \geq 1$ species can unambiguously be identified, there exists a collection of non-overlapping (possibly time dependent), open and bounded regions in trait space, i.e., $\{B_i \subset \Omega, i \leq s,  B_i \cap B_j = \emptyset, i \neq j\}$, where the population density is compactly supported and contains only one local maximum. Thus, if we let $\mu_{B_i}^k(n)$, be the $k$'th order moment of $n$ (for $k \in \{0,1,2\}$, \red{and assuming $n$ is smooth enough for these moments to exist}), within the region $B_i$, i.e.,
\bse
\begin{alignat}{1}
\mu_{B_i}^0(n) &:= \int_{B_i} n(x)\dx,\label{m0}\\ 
\mu_{B_i}^1(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)x\dx,\label{m1}\\
\mu_{B_i}^2(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)(x-\mu_{B_i}^1)\otimes (x-\mu_{B_i}^1)\dx,\label{m2}
\end{alignat}
\ese
we can initialize the SLM at $t=t_{0}$ with initial data
\begin{equation}
N_{0} = (\mu_{B_{1}}^0,\mu_{B_{1}}^1,\mu_{B_{1}}^2,\cdots,\mu_{B_{s}}^0,\mu_{B_{s}}^1,\mu_{B_{s}}^2)|_{t=t_0}.
\end{equation}
Solving \eqref{ODEmodel}--\eqref{ODEmodelIC} then amounts to approximating \eqref{m0}--\eqref{m2} for $t \in J$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time discrete PLM}\label{sec:timedisc}
In this section we introduce the time discrete PLM, which will serve as the basis for the a posteriori error bounds. First, we establish the existence of a solution for every discrete time, then we derive an identity for the residual which also provides uniqueness of the time-discrete solution.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}
For $k = 0,\cdots,M$, we let $t_k$ denote the discrete times, such that $0 = t_0 < t_1 < \cdots < t_M := T$, and where $t_k = t_{k-1} + \tau_k = \sum_{j = 1}^k \tau_k$, for some set of time increments $\{\tau_k\}_{k=1}^M$. In the following, we use superscripts to indicate dependency on the discrete times, e.g., $u^k := u(t_k)$.

@@

$\psi$ satisfies~\eqref{weak_form}, which implies that $\vertiii{\mathcal{R}^k(\psi)}_{k^\ast}=0$. \red{Then, thanks to the identity~\eqref{res_error}, and since $\vertiii{\Rcal^k(\psi)}^2_{k^\ast}$ is convex due to the monotonicity condition~\eqref{monotonicity}, we obtain that $\psi=n^{k}$ in $X_k$.} This section concludes the proof of Theorem~\ref{thm:wellposed}.

@@

We assume the discretization errors associated with solving the models are negligible compared to the modeling error which we are interested in, hence in practice we calculate only the remainder estimator $\eta_{\textn{rem},i}^k$ from \eqref{etarem} when estimating~\eqref{finalestimate} \red{(note that we do not normalize in the plots)}. Moreover, the density reconstruction operator $\Dcal$ is implemented as a normal distribution, i.e., for a species $N_i^k = (n_i^k,x_i^k,\upsilon_i^k)$ we have

@@

With the parameters given in the previous section, we employ both the heuristic and multi-scale speciation algorithms. \red{Figure~\ref{ex1:abundance} below shows the species abundance, figure~\ref{ex1:x1} the species mean trait coordinates, figure~\ref{ex1:variance} the largest eigenvalue of the trait covariance matrix, figure~\ref{ex1:residuals} the a posteriori error bound, and figure~\ref{ex1:edist} the error in energy norm, as functions of time.}

>>>>

Find $N :(0,T) \rightarrow \real^{s(1+d+d^2)}$ such that $n_i \geq 0$, for $1\leq i \leq s$, and 
\dfrac{\textnormal{d}N}{\textnormal{d}t} &= \Ncal(N), \quad && t\in(0,T), \label{ODEmodel}\\
N &= N_0, &&t = 0. \label{ODEmodelIC}
In practice, the problem \eqref{ODEmodel}--\eqref{ODEmodelIC} can be reduced to $s(1+d+d(d+1)/2)$ equations due to the symmetry of the trait covariance matrix. For additional details including a derivation of the above model, see~\cite{nordbotten2020dynamics}. Note also that when indexing the components of a point $x\in\Omega$ we use subscripts, i.e., $x = (x_1,\cdots,x_d)$, not to be confused with the species mean trait coordinate $x_i$, where we write $x_i(t) = (x_{i,1}(t),\cdots,x_{i,d}(t))$. 
%The distinction should be clear from the context.
\subsection{Relationship between population and species level models}
If there is a time interval $J := (t_{0},t_1) \subset (0,T)$ where $s \geq 1$ species can unambiguously be identified, there exists a collection of non-overlapping (possibly time dependent), open and bounded regions in trait space, i.e., $\{B_i \subset \Omega, i \leq s,  B_i \cap B_j = \emptyset, i \neq j\}$, where the population density is compactly supported and contains only one local maximum. Thus, if we let $\mu_{B_i}^k(n)$, be the $k$'th order moment of $n$ (for $k \in \{0,1,2\}$, \red{and assuming $n$ is smooth enough for these moments to exist}), within the region $B_i$, i.e.,
\begin{alignat}{1}
\mu_{B_i}^0(n) &:= \int_{B_i} n(x)\dx,\label{m0}\\ 
\mu_{B_i}^1(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)x\dx,\label{m1}\\
\mu_{B_i}^2(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)(x-\mu_{B_i}^1)\otimes (x-\mu_{B_i}^1)\dx,\label{m2}
we can initialize the SLM at $t=t_{0}$ with initial data
N_{0} = (\mu_{B_{1}}^0,\mu_{B_{1}}^1,\mu_{B_{1}}^2,\cdots,\mu_{B_{s}}^0,\mu_{B_{s}}^1,\mu_{B_{s}}^2)|_{t=t_0}.
Solving \eqref{ODEmodel}--\eqref{ODEmodelIC} then amounts to approximating \eqref{m0}--\eqref{m2} for $t \in J$.
\section{Time discrete PLM}\label{sec:timedisc}
In this section we introduce the time discrete PLM, which will serve as the basis for the a posteriori error bounds: 
\red{
For $k = 0,\cdots,M$, we let $t_k$ denote the discrete times, such that $0 = t_0 < t_1 < \cdots < t_M := T$, and where $t_k = t_{k-1} + \tau_k = \sum_{j = 1}^k \tau_k$, for some set of time increments $\{\tau_k\}_{k=1}^M$ (in the following, we use superscripts to indicate dependency on the discrete times, e.g., $u^k := u(t_k)$). The time-discrete version of eq.~\ref{PDE} then reads as (for $k\geq 1$)
n^k - r^k n^k + \Phi^k(n^k)n^k - \nabla \cdot (g^k \nabla n^k) = f + n^{k-1},\label{timediscPDE}
where we have discretized in time using a standard implicit first order method (i.e., the backward Euler method). While we do not prove convergence of the above discrete scheme to the original continuous problem, the truncation error for the backward Euler method is known to be of first order. In the following, we first establish the existence of a solution to \eqref{timediscPDE} in the weak sense for every discrete time, then we derive an identity for the residual which also provides uniqueness of the time-discrete solution.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}

@@

$\psi$ satisfies~\eqref{weak_form}, which implies that $\vertiii{\mathcal{R}^k(\psi)}_{k^\ast}=0$. \red{Then, since the first term on the right hand side of~\eqref{res_error} is a norm and since the second term is convex due to the monotonicity condition~\eqref{monotonicity}, we obtain that $\psi=n^{k}$ in $X_k$.} This section concludes the proof of Theorem~\ref{thm:wellposed}.

@@

We assume the discretization errors associated with solving the models are negligible compared to the modeling error which we are interested in, hence in practice we calculate only the remainder estimator $\eta_{\textn{rem},i}^k$ from \eqref{etarem} when estimating~\eqref{finalestimate}. Moreover, the density reconstruction operator $\Dcal$ is implemented as a normal distribution, i.e., for a species $N_i^k = (n_i^k,x_i^k,\upsilon_i^k)$ we have

@@

With the parameters given in the previous section, we employ both the heuristic and multi-scale speciation algorithms. \red{Figure~\ref{ex1:abundance} below shows the species abundance, figure~\ref{ex1:x1} the species mean trait coordinates, figure~\ref{ex1:variance} the largest eigenvalue of the trait covariance matrix, figure~\ref{ex1:residuals} the a posteriori error bound, and figure~\ref{ex1:edist} the error in energy norm, as functions of time. Note that due to the presence of $G^{-\frac{1}{2}}$ in the definition of $\eta_{rem,i}^k$ (eq. \eqref{etarem}), the magnitudes in figure~\ref{ex1:residuals} can not be directly compared to those of the preceding figures.}
